Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2023-01-24 tensor_clustering/IEEE/ieee_revision/response_s1.tex

\item Comparison of assumptions with previous work. Our constraints in $\tP$ are mild compared with previous literature. We have also added following comparison in Section~\ref{subsec:identify}:
\begin{quote}
    \begin{table}[h]
    \centering
    \resizebox{\textwidth}{!}{
    \begin{tabular}{c|c ccc}
    \hline
    Assumptions in parameter space& \cite{gao2018community}& \cite{han2020exact}& \cite{ke2019community} &Ours\\
    \hline
         Balanced community sizes &$\surd$ &$\surd$&$\surd$ & $\surd$  \\
         Bounded core tensors& $\surd$ & $\times$ & $\surd$ &$\surd$\\
              Balanced degrees & $\surd$&-&$\surd$ &$\surd$\\
         Flexible in-group connections &$\times$ &$\surd$&$\surd$& $\surd$\\
         Gaps among cluster centers & In-between cluster difference & Euclidean gap & Eigen gap & Angle gap\\
         \hline
    \end{tabular}
    }
    \caption{Parameter space comparison between previous work with our assumption.}
\end{table}
\end{quote}

% We agree that our dTBM is identifiable with some degenerated rank, and we thank the reviewer for pointing out the incorrect expression. The upper bound is technical constraint for theoretical analysis; in practice, the constraint $\max_{a:} \onorm{\mat(\mS)_{a:}} \leq c_4$ would likely never be active with a large $c_4 \geq \onorm{\tY}$.  Nevertheless, the lower bound of core tensor $\tS$ is still necessary for degree-corrected models. The lower bound is critical for identifiability; a purely zero slice in $\tS$ makes the heterogeneity indistinguishable. We have added a short counter-example to illustrate the necessity of non-zero core tensor slice.

% {\color{blue}

% \begin{example}[Non-identifiability with pure zero core slice] Consider an order-2 dTBM with core tensor $\mS = \begin{pmatrix} 0 & 0\\
%     1 & -1
%     \end{pmatrix}$ and mean tensor 
%     \begin{equation}
%         \tX =  \mTheta_1 \mM \mS   \mM^T \mTheta_2, \quad \text{with } \mM = \begin{bmatrix} 1 & 0 \\
%         1 & 0 \\
%         0 & 1 \\
%         0 & 1 \\
%         \end{bmatrix}, \quad \mTheta_1 =\mTheta_2= \text{diag}(1,1,1,1).
%     \end{equation}
% Replacing $\mTheta_1$ by $\mTheta_1 = (3/2, 1/2, 1,1)$ leads to the same mean tensor $\tX$.
% \end{example}
% }

% In addition, the constraints in our parameter space are mild compared with previous degree-corrected block models \citep{gao2018community, ke2019community}. We have added a parameter space comparison table.

% {
% \color{blue}

% \begin{table}[h]
%     \centering
%     \resizebox{\textwidth}{!}{
%     \begin{tabular}{c|c ccc}
%     \hline
%     Assumptions in parameter space& \cite{gao2018community}& \cite{han2020exact}& \cite{ke2019community} &Ours\\
%     \hline
%          Balanced community size &$\surd$ &$\surd$&$\surd$ & $\surd$  \\
%          Balanced degree parameters & $\surd$&-&$\surd$ &$\surd$\\
%          Bounded connection slices & $\surd$ & $\times$ & $\surd$ &$\surd$\\
%          Flexible in-group connections &$\times$ &$\surd$&$\times$& $\surd$\\
%          Gap among cluster centers & In-between cluster difference & Euclidean gap & Eigen gap & Angle gap\\
%          \hline
%     \end{tabular}
%     }
%     \caption{Parameter space comparison between previous work with our assumption.}
% \end{table}
% }

\end{enumerate}
\item \textit{Another big concern I have is the whole Section 3. The statistical and computational lower bounds are exactly the same as the tensor block model in \cite{han2020exact}. Therefore, the whole results in Section 3 can be simply obtained as corollary from \cite{han2020exact}. since the authors are working on a more restrictive model and the lower bounds must be larger than simpler model.}

\textbf{Response:}

In fact, none of our results in Section 3 can be derived from earlier non-degree work~\citep{han2020exact}. The short answer is that, our results show the similar conclusion but {\bf under different conditions.} While the TBM impossibility~\citep{han2020exact} provides a necessary condition for our dTBM impossibility, we find that {\bf such a condition is loose} in our contexts. This has motivated to develop new techniques to derive sharp lower bounds. 
Inspired by your comment, we have revised the exposition of our theory. See our response to \textbf{\# 1 Major Comments from Editor} for details.

We quote the newly-added texts here:
\begin{quote}
    ``...\emph{Theoretical results.} In both works, we study the phase transition of TBM and dTBM with respect to the Euclidean and angle-based SNRs. We briefly summarize the results in \cite{han2020exact} and compare with ours.

    \textit{Statistical critical value:}
    \begin{align}
        \text{Ours:}& \ \Delta_{\text{ang}}^2 \lesssim p^{-(K-1)} \Rightarrow \text{statistically impossible;} \quad \Delta_{\text{ang}}^2 \gtrsim   p^{-(K-1)} \Rightarrow \text{MLE achieves exact recovery;} \\
        \text{Han's:}& \ \Delta_{\text{Euc}}^2 \lesssim p^{-(K-1)} \Rightarrow \text{statistically impossible;} \quad \Delta_{\text{Euc}}^2 \gtrsim   p^{-(K-1)} \Rightarrow \text{MLE achieves exact recovery}.
    \end{align}

     \textit{Computational critical value:}
    \begin{align}
        \text{Ours:}& \ \Delta_{\text{ang}}^2 \lesssim p^{-K/2} \Rightarrow \text{computationally impossible;} \quad \Delta_{\text{ang}}^2 \gtrsim   p^{-K/2} \Rightarrow \text{ computationally efficient;} \\
        \text{Han's:}& \ \Delta_{\text{Euc}}^2 \lesssim p^{-K/2} \Rightarrow \text{computationally impossible;} \quad \Delta_{\text{Euc}}^2 \gtrsim   p^{-K/2} \Rightarrow \text{computationally efficient}.
    \end{align}

The above comparison reveals three major differences.

 %Our dTBM phase transition re-discovers the statistical-computational gaps in TBM. %This finding indicates the intrinsic behavior distinctions among higher-order, matrix, and vector problem may be generally exists in many higher-order problems.
First, none of our results in Section~\ref{sec:limits} are corollaries of \cite{han2020exact}. Both models show the similar conclusion {\bf but under different conditions}. While the TBM impossibility~\citep{han2020exact} provides a necessary condition for our dTBM impossibility, we find that such a condition is often loose. There exists a regime of $\tS$ in which TBM problems are computationally efficient but dTBM problems are statistically impossible; see Example~\ref{example:euc_alg}. This observation has motivated us to develop the new signal notion $\Delta^2_{\text{ang}}$ for sharp dTBM phase transition conditions.

 Second, to find the phase transition, we need to show both the impossibility {\bf and} achievability when SNR is below {\bf and} above the critical value, respectively. While the TBM impossibility can serve as a loose condition of our dTBM impossibility, more efforts are required to show the achievability. In particular, since TBM is a more restrictive model than dTBM, the achievability in \cite{han2020exact} does not imply the achievability of dTBM in a larger parameter space. The latter requires us to develop new MLE and polynomial algorithms for dTBM achievability.  %For impossibility, the minimax bounds in dTBM search over a larger range of estimators than TBM. Though the thresholds for TBM impossibility can serve as the loose lower bounds of the dTBM impossibility, more efforts are required to avoid the gap between impossibility and achievablity.

Third, from the perspective of proofs, we develop new tools to handle the extra degree heterogeneity. In our Theorem~\ref{thm:stats}, we consider the profile MLE by maximizing out the nuisance degree parameter, while TBM~\citep{han2020exact} considers the usual MLE without degree parameter. In our Theorem~\ref{thm:comp}, we construct a rank-2 tensor to relate HPC conjecture to $\Delta^2_{\text{ang}}$, while TBM~\citep{han2020exact} constructs a rank-1 tensor to relate HPC conjecture to $\Delta^2_{\text{Euc}}$. The asymptotic non-equivalence between $\Delta^2_{\text{ang}}$ and $\Delta^2_{\text{Euc}}$ renders our proof technically more involved...''
   \end{quote}

   \begin{quote}
   ``... \emph{Signal notion.} We emphasize that the signal notions are different between the two models. In particular, the Euclidean-based signal notion in TBM~\cite{han2020exact} fails to accurately describe the phase transition in our dTBM due to the possible heterogeneity in degree $\mtheta$. To compare, we denote our angle-based signal notion in~\eqref{eq:gamma} and the Euclidean-based SNR in \cite{han2020exact} as $\Delta_{\text{ang}}^2$ and $\Delta_{\text{Euc}}^2$, respectively:
\begin{equation}
     \Delta_{\text{ang}}^2 =  2(1 - \max_{a \neq b\in [r]}\cos \of{\mS_{a:},\  \mS_{b:}} ), \quad \Delta_{\text{Euc}}^2 = \min_{a \neq b \in [r]} \onormSize{}{\mS_{a:} - \mS_{b:}}^2.
\end{equation}
By Lemma~\ref{lem:norm_diff} in the Appendix II, we have 
\begin{equation}\label{eq:signalcompare}
     \Delta_{\text{ang}}^2  \max_{a \in [r]}\onormSize{}{\mS_{a:}}^2 \leq \Delta_{\text{Euc}}^2.
\end{equation}
The above inequality indicates the condition $\Delta_{\text{Euc}}^2 \leq p^{\gamma}$ is {\bf sufficient but not necessary} for $\Delta_{\text{ang}}^2 \leq p^{\gamma}$. In fact, if we were to use $\Delta_{\text{Euc}}^2$ for both models, then the phase transition of dTBM can be arbitrarily worse than that for TBM.

Here, we provide an example to illustrate the dramatical difference between TBM and dTBM with the same core tensor.

\begin{example}[Comparison with Euclidean-based signal notion] \label{example:euc_alg} Consider a biclustering model with $\mtheta=1$ and an order-2 core matrix 
\begin{equation}
    \mS = \begin{pmatrix} p^{(\gamma+1)/2 } + 2  & 2 p^{(\gamma+1)/2} + 4\\
    2 & 4
    \end{pmatrix},\quad \text{with}\ \gamma \leq -1.
\end{equation}
The core matrix $\mS$ lies in the parameter spaces of TBM and our dTBM. Here, the constraint $\gamma \leq -1$ is added to ensure the bounded condition of $\mS$ in our parameter space in \eqref{eq:family}. The angle-based and Euclidean-based signal levels of $\mS$ are 
\begin{equation}
    \Delta_{\text{ang }}^2(\mS) = 0 \ \left(\leq p^{\gamma}\right), \quad \Delta_{\text{Euc}}^2(\mS) = 5 p^{\gamma + 1} \ \left(\geq p^{\gamma}\right).
\end{equation}
We conclude that TBM with $\mS$ achieves exact recovery with a polynomial-time algorithm; see \citet[Theorem 4]{han2020exact}. By contrast, the dTBM with the same $\mS$ and input $r=2$ violets the identifiability condition, and thus fails to be solved by all estimators; see our Theorem~\ref{thm:unique}.'' 
\end{example}

   \end{quote}
   We have revised our Section~\ref{sec:limits} to better present the lower bounds. See our response to \textbf{\# 1 Major Comments from Editor}.

\newpage

\begin{algorithm}[h!]
\caption*{\bf Algorithm: Multiway spherical clustering for degree-corrected tensor block model }
\vspace{.15cm}
\begin{algorithmic}[1]
\Algphase{Sub-algorithm 1: Weighted higher-order initialization}
\INPUT Observation $\tY \in \bbR^{p\times \cdots \times p}$, cluster number $r$, relaxation factor $\eta > 1$ in $k$-means clustering.

\State {
\color{blue}Compute factor matrix $ \mU_{\text{pre},k} = \text{SVD}_{r} (\text{Mat}_k(\tY)), k \in [K]$ and the $(K-1)$-mode projection 
\begin{equation}
    \tX_{\text{pre},k} = \tY \times_1   \mU_{\text{pre},1} \mU_{\text{pre},1}^T \times_2 \cdots  \times_{k-1} \mU_{\text{pre},k-1} \mU_{\text{pre},k-1}^T \times_{k+1} \mU_{\text{pre},k-1} \mU_{\text{pre},k+1}^T\times_{K+1}  \times_K \mU_{\text{pre},K} \mU_{\text{pre}, K}^T.
\end{equation}
}
\State {
\color{blue}
Compute factor matrix $\hat \mU_k = \text{SVD}_{r}(\text{Mat}_k(\tX_{\text{pre},k})), k \in [K]$ and denoised tensor
\begin{equation}
    \hat \tX = \tY \times_1 \hat \mU_1 \hat \mU^T_1 \times_2 \cdots \times_K \hat \mU_K \hat \mU^T_K.
\end{equation}
} 
\For{\textcolor{blue}{$k \in [K]$}}
\State { \color{blue} Let $\hat \mX = \text{Mat}_k(\hat \tX)$ and $S_0=\{i \in [p]: \onormSize{}{\hat \mX_{i:}} = 0\}$. Set $\hat z(i)$ randomly in $[r]$ for $i \in S_0$.}
\State{  For all $i\in S_0^c$, compute normalized rows
$\hat \mX_{i:}^s :=\onormSize{}{\hat \mX_{i:}}^{-1} \hat \mX_{i:}.$
}
\State { Solve the clustering $\hat z_k \colon [p]\to[r]$ and centroids $ (\hat \mx_j)_{j\in[r_k]}$ using weighted $k$-means, such that}
\begin{align}
    &\sum_{i \in S_0^c }  \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{\hat \mX_{i:}^s - \hat \mx_{\hat z_k(i)} }^2 
    \leq 
    \eta \min_{\substack{\bar \mx_j, j\in[r], \bar z_k(i),i\in S_0^{c}}} \sum_{i \in S^c } \onormSize{}{\hat \mX_{i:}}^2 \onormSize{}{ \hat \mX_{i:}^s -   \bar \mx_{\hat z_k(i)}}^2.
\end{align}
\EndFor

\OUTPUT {\color{blue} Initial clustering $z^{(0)}_k \leftarrow \hat z_k, k \in [K]$.}

\Algphase{Sub-algorithm 2: Angle-based iteration}
\INPUT Observation $\tY \in \bbR^{p \times \cdots \times p}$, initialization $z^{(0)}_k \colon [p]\to[r], k \in [K]$ from Sub-algorithm 1, iteration number $T$.
\For {$t = 0$ to $T-1$}
\State Update the block tensor $\tS^{(t)}$ via
$\tS^{(t)} (a_1,...,a_K)= \text{Ave} \{\tY(i_1,\ldots,i_K): z^{(t)}_k(i_k) = a_k, k \in [K]\}.$
\For{ $k \in [K]$}
\State {\color{blue} Calculate reduced tensor $\tY^{\text{d}}_k \in \bbR^{r \times r \times p \times r \times \cdots \times r}$ via
\begin{equation}
    \tY^{\text{d}}_k(a_1,\ldots,a_{k-1}, i ,a_{k+1},\ldots,a_K) 
    = \text{Ave}\{\tY(i_1,\ldots,i_{k-1},i,i_{k+1},\ldots,i_K): z^{(t)}(i_j) = a_j, j \neq k \}
\end{equation}}
% \begin{align}
%     &\tY^{\text{d}}(i,a_2,\ldots,a_K) 
%     = 
% \text{Ave}\{\tY(i,i_2,\ldots,i_K): z^{(t)}(i_k) = a_k, k \neq 1 \}.
% \end{align}

\State { \color{blue} Let $\mY_k^{\text{d}} = \text{Mat}_k(\tY^{\text{d}})$ and $J_0 = \{ i\in[p]: \onorm{\mY^{\text{d}}_{i:}} = 0\}$. Set $z_k^{(t+1)}(i)$ randomly in $[r]$ for $i \in J_0$.}

\State {\color{blue} Let $\mS^{(t)} = \text{Mat}(\tS^{(t)})$. For all $i \in J_0^c$ update the cluster assignment by
\begin{equation}
    z(i)^{(t+1)}_k = \argmax_{a \in [r]} \cos \left( \mY^{\text{d}}_{k,i:},\ \mS^{(t)}_{a:} \right).
\end{equation}}
\EndFor
\EndFor

\OUTPUT { \color{blue} Estimated clustering $z^{(T)}_k  \in [r]^{p}, k \in [K]$.}

\end{algorithmic}
\end{algorithm}

\end{enumerate}

% \begin{lem}[perturbation for degree-corrected clustering]. Let $z^*$ be the true clustering. Fix an arbitrary clustering $z$ with $\text{MisClust}(z,z^*):=\varepsilon \leq ...$ (a small constant). 
% Under assumptions....., we have
% \begin{equation}\label{eq:mode1}
% \min_{(\tS, \Theta)}\Fnorm{\tX(z^*, \tS^*, \Theta^*)-\tX(z; \tS, \Theta)}^2\gtrsim (...\text{factors involving $p^K$... }) \varepsilon \text{Angle-gap}^2 + o(\varepsilon).
% \end{equation}
% \end{lem}

% \begin{rmk}
% Everything is deterministic. Here, $\tX(z^*,...)$ is the degree-corrected signal tensor. Two ways to prove: 1) Find (deterministic)
% \[
% (\tS_{\min}, \Theta_{\min})=\argmin_{(\tS, \Theta)}\Fnorm{\tX(z^*)-\tX(z, \tS, \Theta)}^2,
% \]
% and then plug into~\eqref{eq:mode1} to evaluate the perturbation; 2) Prove a stronger version of identifiability result
% \[
% \KeepStyleUnderBrace{\min_{(\tS, \Theta, \tS^*, \Theta^*)}\Fnorm{\tX(z^*, \tS^*, \Theta^*)-\tX(z; \tS, \Theta)}^2}_{\text{distance between two points in the parameter space}}\gtrsim (...\text{factors involving $p^K$... })  \text{Angle-gap}^2 \KeepStyleUnderBrace{\varepsilon}_{\text{distance between $z$ and $z^*$}}.
% \]
% Our identifiability theorem is a special case of the above inequality by right hand side = 0.

### 2023-01-24 tensor_clustering/IEEE/ieee_revision/response_s1.tex

% Lemma 1 is simpler than approaches in my NIPS paper, because no $\tY$ or expectation is needed. 
% In addition, I omitted the subscript $k$ in~\eqref{eq:mode1} for intuition. Precisely, ~\eqref{eq:mode1} holds only for $z_1$ while fixing $\{z_k\}_{k\geq 2}$. 
% \end{rmk}
% \begin{lem}[Union bound]
% \begin{align}
% &\mathbb{P}(\text{MisClust}(\hat z_{\text{MLE}}, z^*)\geq \varepsilon)\\
% \stackrel{\text{\color{red}You should write this step as instinct}}{=}&\mathbb{P}(\exists (z, \tS, \Theta) \ \text{s.t.}\  \text{MisClust}(z,z^*)\geq \varepsilon \ \text{and}\ \text{Log-lik}(z, \tS, \Theta) \geq \text{Log-lik}(z^*))\\
% %&\leq \sum_{z \text{ for which MisClust}(z,z^*)\geq \varepsilon}\mathbb{P}(\text{Log-lik}(z)\geq \text{Log-lik}(z^*))\\
% \leq &\sum_{z \text{ for which MisClust}(z,z^*)\geq \varepsilon}\sum_{(\tS, \Theta)}\mathbb{P}(\Fnorm{\tY-\tX(z, \tS, \Theta)}^2\leq \Fnorm{\tY-\tX(z^*)}^2)\\
% \leq & \sum_{z:\ \text{MisClust}(z,z^*)\geq \varepsilon}\sum_{(\tS, \Theta)} \mathbb{P}\left({\langle \tE, \tX(z^*)-\tX(z, \tS, \Theta)\rangle  \over\Fnorm{\tX(z^*)-\tX(z, \tS, \Theta)}}\gtrsim \Fnorm{\tX(z^*)-\tX(z, \tS, \Theta)} \right)\\
% \leq& \text{(desired conclusion)}
% \end{align}

% In the last step, only randomness is on $\tE$. The term $\Fnorm{\tX(z^*)-\tX(z,...)}$ can be replaced by $...\sqrt{\varepsilon}$ because of condition in the sum ``$z$ for which MisClust$(z,z^*)\geq \varepsilon$ '' and Lemma 1.

% \fixme{Miaoyan}{Please complete full steps using ideas from Lemma 1 and Thm 1 in my NIPS paper. Lemma 2 is simple. I sketched out the proof for Lemma 1 for $R=2$. For general $R$ (number of blocks), it should be doable (I did not do) with just more notations}

% \end{lem}

\bibliography{tensor_wang}
\bibliographystyle{apalike}

\end{document}
